Comparison of large-scale citizen science data and long-term study data for phenology modeling

### 2018-09-12 manuscript/supplement_images.tex
final revisions edits from shawn before sending to co-auths

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{lineno}
\usepackage{setspace}
  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}
{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}

@@

Supplementary images S1 - S7 and Table S1

@@

\textbf{Figure S1}: RMSE for specific species and phenophases of the NPN dataset. X marks the best performing models for each respective data type.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_npn_models.png}
	Figure S1
\end{center}

\newpage
%%%%%%%%%%%%%%%%%
%% Figure S2
%%%%%%%%%%%%%%%%%
\textbf{Figure S2}: RMSE for specific species and phenophases of the LTER datasets. X marks the best performing models for each respective data type.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_best_lter_models.png}
	Figure S2
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S3
%%%%%%%%%%%%%%%%%

@@

\textbf{Figure S3}: RMSE of all species and phenophases of the four scenarios described in the text. These values were calculated using held out test data.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
	Figure S3
\end{center}

@@

\textbf{Figure S4}: Distribution of parameters of the Naive, GDD, Fixed GDD, and Linear models for the three species common to the Hubbard Brook, Harvard, and NPN datasets. The phenophase is budburst for all three species. Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution. Note the heading for each sub figure.

@@

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
	Figure S4
\end{center}

@@

\textbf{Figure S5}: As in Figure S4, but for the Alternating and Uniforc models.

@@

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
	Figure S5
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S6
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S6}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and NPN derived models. The phenophase for the four species is budburst. These 4 species are representative of the analysis, and for the remaining comparisons the reader is pointed to the script 'analysis/plot\_select\_species\_parameters.R' in the code repository to generate additional figures.

@@

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
	Figure S6
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S7
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S7}: As in Figure S6, but for the Alternating and Uniforc models.

@@

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
	Figure S7
\end{center}
%%%%%%%%%%%%%%%%%
%% Table S1
%%%%%%%%%%%%%%%%%
\newpage

@@

\begin{table}
\tiny
\begin{tabular}{l|r|l|l|l|l|l|l}
\hline
species & phenophase & phenophase\_type & harvard & hjandrews & hubbard & jornada & npn\\
\hline
acer circinatum & 371 & Budburst & - & 266 (66) & - & - & 39 (10)\\
\hline
acer circinatum & 501 & Flowers & - & 116 (29) & - & - & 33 (8)\\
\hline
acer pensylvanicum & 371 & Budburst & 80 (20) & - & - & - & 34 (9)\\
\hline
acer rubrum & 371 & Budburst & 100 (25) & - & - & - & 957 (239)\\
\hline
acer rubrum & 501 & Flowers & 96 (24) & - & - & - & 668 (167)\\
\hline
acer saccharum & 371 & Budburst & 60 (15) & - & 164 (41) & - & 365 (91)\\
\hline
betula alleghaniensis & 371 & Budburst & 60 (15) & - & 178 (44) & - & 133 (33)\\
\hline
betula alleghaniensis & 501 & Flowers & 26 (7) & - & - & - & 64 (16)\\
\hline
betula lenta & 371 & Budburst & 58 (14) & - & - & - & 96 (24)\\
\hline
betula papyrifera & 371 & Budburst & 76 (19) & - & - & - & 96 (26)\\
\hline
betula papyrifera & 501 & Flowers & 18 (5) & - & - & - & 36 (9)\\
\hline
fagus grandifolia & 371 & Budburst & 76 (19) & - & 177 (44) & - & 259 (65)\\
\hline
fraxinus americana & 371 & Budburst & 84 (21) & - & - & - & 90 (23)\\
\hline
fraxinus americana & 501 & Flowers & 22 (6) & - & - & - & 52 (13)\\
\hline
ilex verticillata & 371 & Budburst & 35 (9) & - & - & - & 26 (6)\\
\hline
larrea tridentata & 501 & Flowers & - & - & - & 27 (7) & 118 (30)\\
\hline
nyssa sylvatica & 371 & Budburst & 27 (7) & - & - & - & 63 (16)\\
\hline
pinus strobus & 496 & Budburst & 38 (10) & - & - & - & 77 (19)\\
\hline
populus tremuloides & 371 & Budburst & 38 (10) & - & - & - & 208 (51)\\
\hline
populus tremuloides & 501 & Flowers & 17 (4) & - & - & - & 79 (22)\\
\hline
prosopis glandulosa & 501 & Flowers & - & - & - & 49 (12) & 78 (20)\\
\hline
prunus serotina & 371 & Budburst & 58 (14) & - & - & - & 228 (57)\\
\hline
pseudotsuga menziesii & 480 & Budburst & - & 182 (46) & - & - & 38 (10)\\
\hline
quercus alba & 371 & Budburst & 62 (15) & - & - & - & 174 (43)\\
\hline
quercus rubra & 371 & Budburst & 80 (20) & - & - & - & 242 (60)\\
\hline
quercus rubra & 501 & Flowers & 56 (14) & - & - & - & 127 (32)\\
\hline
quercus velutina & 371 & Budburst & 77 (19) & - & - & - & 72 (18)\\
\hline
rhododendron macrophyllum & 371 & Budburst & - & 84 (21) & - & - & 48 (12)\\
\hline
rhododendron macrophyllum & 501 & Flowers & - & 27 (7) & - & - & 50 (12)\\
\hline
trillium ovatum & 488 & Budburst & - & 222 (55) & - & - & 68 (17)\\
\hline
trillium ovatum & 501 & Flowers & - & 169 (42) & - & - & 60 (15)\\
\hline
vaccinium corymbosum & 371 & Budburst & 38 (10) & - & - & - & 60 (15)\\
\hline
vaccinium corymbosum & 501 & Flowers & 38 (10) & - & - & - & 65 (16)\\
\hline
vaccinium parvifolium & 371 & Budburst & - & 149 (37) & - & - & 25 (6)\\
\hline
vaccinium parvifolium & 501 & Flowers & - & 162 (41) & - & - & 25 (6)\\
\hline
\end{tabular}
\newline
Table S1
\end{table}

>>>>

\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{lineno}
\usepackage{setspace}
  \doublespacing
\usepackage[document]{ragged2e}

\begin{document}
{\Comparison of large-scale citizen science data and long-term study data for phenology modeling \par}

@@

Supplementary images S1 - S11 and Table S1

@@

\textbf{Figure S1}: Sensitivity test results from using a 15 versus 30 day cutoff between the 'yes' and most recent 'no' in the USA-NPN dataset. Each point represents the value of a single parameter for one of eight models for 33 species and phenophases in the USA-NPN dataset. Note only 33 comparisons were possible here as the stricter 15 day cutoff resulted in three of the species not having sufficient observations.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{parameter_estimates_from_cutoff_sensitivity.png}
	Figure S1
\end{center}

@@

%%%%%%%%%%%%%%%%%
%% Figure S2
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S2}: Comparisons of parameter estimates between NPN and LTER derived models. As in Figure 2 in the main text, but using a threshold of 15 instead of 30 days between the first 'yes' and most recent 'no' in the USA-NPN dataset. See methods for details.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{figure_param_comparison_15day_cutoff.png}
	Figure S2
\end{center}

@@

%%%%%%%%%%%%%%%%%
%% Figure S3
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S3}: Comparison of predicted day of year (DOY) of all phenological events between NPN
and LTER-derived models. As in Figure 3 in the main text, but using a threshold of 15 instead of 30 days between the first 'yes' and most recent 'no' in the USA-NPN dataset. See methods for details.

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{figure_estimate_compare_15day_cutoff.png}
	Figure S3
\end{center}

@@

\textbf{Figure S4}: Differences in prediction error between NPN and LTER-derived models. As in Figure 4 in the main text, but using a threshold of 15 instead of 30 days between the first 'yes' and most recent 'no' in the USA-NPN dataset. See methods for details.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{figure_rmse_metrics_density_plot_15day_cutoff.png}
	Figure S4
\end{center}

@@

\textbf{Figure S5}: RMSE for specific species and phenophases using all combinations of models and data sources. Red X's mark the best performing models for each respective dataset.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_all_model_rmse.png}
	Figure S5
\end{center}

\newpage
%%%%%%%%%%%%%%%%%
%% Figure S6
%%%%%%%%%%%%%%%%%
\textbf{Figure S6}: Pearson correlation coefficients for specific species and phenophases using all combinations of models and data sources. Note that since all predictions from each Naive model are the same pearson's correlation cannot be calculated here. Red X's mark the best performing models for each respective dataset.

@@

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_all_model_pearson.png}
	Figure S6
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S7
%%%%%%%%%%%%%%%%%

@@

\textbf{Figure S7}: RMSE of all species and phenophases of the four scenarios described in the text. These values were calculated using held out test data.

\begin{center}
	\centering
		\includegraphics[width=1\textwidth]{supplement_scenario_absolute_rmse.png}
	Figure S7
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S8
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S8}: Distribution of parameters of the Naive, GDD, Fixed GDD, and Linear models for the three species common to the Hubbard Brook, Harvard, and USA-NPN datasets. The phenophase is budburst for all three species. Vertical lines indicate either the mean (solid) or median (dashed) of the respective distribution. Note the heading for each sub figure.

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison1.png}
	Figure S8
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S9
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S9}: As in Figure S4, but for the Alternating and Uniforc models.

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_hubbard_harvard_comparison2.png}
	Figure S9
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S10
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S10}: As in Figure S4, but for 4 selected species to show the difference in parameter distributions between LTER and USA-NPN derived models. The phenophase for the four species is budburst. These 4 species are representative of the analysis, and for the remaining comparisons the reader is pointed to the script 'analysis/plot\_select\_species\_parameters.R' in the code repository to generate additional figures.

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_select_species_param_comparison1.png}
	Figure S10
\end{center}

%%%%%%%%%%%%%%%%%
%% Figure S11
%%%%%%%%%%%%%%%%%
\newpage

\textbf{Figure S11}: As in Figure S6, but for the Alternating and Uniforc models.

\begin{center}
	\centering
		\includegraphics[scale=0.5]{supplement_select_species_param_comparison2.png}
	Figure S11
\end{center}
%%%%%%%%%%%%%%%%%
%% Table S1
%%%%%%%%%%%%%%%%%
\newpage

@@

\tiny
\begin{tabular}{l|r|l|l|l|l|l|l}
\hline
species & phenophase & phenophase\_type & harvard & hjandrews & hubbard & jornada & npn\\
\hline
acer circinatum & 371 & Budburst & - & 266 (66) & - & - & 39 (10)\\
\hline
acer circinatum & 501 & Flowers & - & 116 (29) & - & - & 33 (8)\\
\hline
acer pensylvanicum & 371 & Budburst & 80 (20) & - & - & - & 34 (9)\\
\hline
acer rubrum & 371 & Budburst & 100 (25) & - & - & - & 957 (239)\\
\hline
acer rubrum & 501 & Flowers & 96 (24) & - & - & - & 668 (167)\\
\hline
acer saccharum & 371 & Budburst & 60 (15) & - & 164 (41) & - & 365 (91)\\
\hline
betula alleghaniensis & 371 & Budburst & 60 (15) & - & 178 (44) & - & 133 (33)\\
\hline
betula alleghaniensis & 501 & Flowers & 26 (7) & - & - & - & 64 (16)\\
\hline
betula lenta & 371 & Budburst & 58 (14) & - & - & - & 96 (24)\\
\hline
betula papyrifera & 371 & Budburst & 76 (19) & - & - & - & 96 (26)\\
\hline
betula papyrifera & 501 & Flowers & 18 (5) & - & - & - & 36 (9)\\
\hline
fagus grandifolia & 371 & Budburst & 76 (19) & - & 177 (44) & - & 259 (65)\\
\hline
fraxinus americana & 371 & Budburst & 84 (21) & - & - & - & 90 (23)\\
\hline
fraxinus americana & 501 & Flowers & 22 (6) & - & - & - & 52 (13)\\
\hline
ilex verticillata & 371 & Budburst & 35 (9) & - & - & - & 26 (6)\\
\hline
larrea tridentata & 501 & Flowers & - & - & - & 27 (7) & 118 (30)\\
\hline
nyssa sylvatica & 371 & Budburst & 27 (7) & - & - & - & 63 (16)\\
\hline
pinus strobus & 496 & Budburst & 38 (10) & - & - & - & 77 (19)\\
\hline
populus tremuloides & 371 & Budburst & 38 (10) & - & - & - & 208 (51)\\
\hline
populus tremuloides & 501 & Flowers & 17 (4) & - & - & - & 79 (22)\\
\hline
prosopis glandulosa & 501 & Flowers & - & - & - & 49 (12) & 78 (20)\\
\hline
prunus serotina & 371 & Budburst & 58 (14) & - & - & - & 228 (57)\\
\hline
pseudotsuga menziesii & 480 & Budburst & - & 182 (46) & - & - & 38 (10)\\
\hline
quercus alba & 371 & Budburst & 62 (15) & - & - & - & 174 (43)\\
\hline
quercus rubra & 371 & Budburst & 80 (20) & - & - & - & 242 (60)\\
\hline
quercus rubra & 501 & Flowers & 56 (14) & - & - & - & 127 (32)\\
\hline
quercus velutina & 371 & Budburst & 77 (19) & - & - & - & 72 (18)\\
\hline
rhododendron macrophyllum & 371 & Budburst & - & 84 (21) & - & - & 48 (12)\\
\hline
rhododendron macrophyllum & 501 & Flowers & - & 27 (7) & - & - & 50 (12)\\
\hline
trillium ovatum & 488 & Budburst & - & 222 (55) & - & - & 68 (17)\\
\hline
trillium ovatum & 501 & Flowers & - & 169 (42) & - & - & 60 (15)\\
\hline
vaccinium corymbosum & 371 & Budburst & 38 (10) & - & - & - & 60 (15)\\
\hline
vaccinium corymbosum & 501 & Flowers & 38 (10) & - & - & - & 65 (16)\\
\hline
vaccinium parvifolium & 371 & Budburst & - & 149 (37) & - & - & 25 (6)\\
\hline
vaccinium parvifolium & 501 & Flowers & - & 162 (41) & - & - & 25 (6)\\
\hline
\end{tabular}
\newline
Table S1 \newline

%%%%%%%%%%%%%%%%%
%% Table S2
%%%%%%%%%%%%%%%%%
\newpage
\normalsize
\textbf{Table S2}: Overall best models when doing cross dataset comparisons. Observations for all species and phenophases were aggregated together to calculate RMSE and Pearsons coefficient for each combination of Parameter source (either USA-NPN or LTER), observation source (either USA-NPN or LTER), and model (6 possible phenology models). Bold indicates the best performing model for a specific parameter and observation combination, with some combinations having ties among multiple models.

\tiny
\begin{adjustbox}{angle=90}
\begin{tabular}{|c|c|cc|cc|cc|cc|cc|cc|cc|cc|}
\hline
\begin{tabular}[c]{@{}c@{}}Parameter\\ Source\end{tabular} & \begin{tabular}[c]{@{}c@{}}Held Out\\ Observation Source\end{tabular} & \multicolumn{2}{c|}{Alternating} & \multicolumn{2}{c|}{Fixed GDD} & \multicolumn{2}{c|}{GDD} & \multicolumn{2}{c|}{Linear} & \multicolumn{2}{c|}{M1} & \multicolumn{2}{c|}{MSB} & \multicolumn{2}{c|}{Naive} & \multicolumn{2}{c|}{Uniforc} \\
 &  & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE & p & RMSE \\ \hline
LTER & LTER & 0.87 & 8.73 & 0.84 & 10.14 & \textbf{0.90} & 7.89 & 0.81 & 10.27 & \textbf{0.90} & 7.89 & 0.87 & 8.73 & 0.72 & 12.35 & \textbf{0.90} & \textbf{7.86} \\ \hline
LTER & USA-NPN & 0.44 & 26.07 & \textbf{0.72} & 22.70 & 0.68 & 20.52 & 0.65 & 22.94 & 0.68 & 20.52 & 0.44 & 26.07 & 0.34 & 31.25 & 0.70 & \textbf{19.69} \\ \hline
USA-NPN & LTER & 0.63 & 16.17 & 0.69 & 16.15 & 0.70 & 13.73 & 0.68 & 16.59 & 0.71 & 13.79 & 0.66 & 15.79 & 0.61 & 27.08 & \textbf{0.72} & \textbf{13.48} \\ \hline
USA-NPN & USA-NPN & 0.80 & 15.27 & 0.76 & 20.18 & \textbf{0.82} & 14.67 & 0.77 & 16.19 & \textbf{0.82} & 14.71 & 0.80 & 15.20 & 0.53 & 21.57 & \textbf{0.82} & \textbf{14.37} \\ \hline
\end{tabular}
\end{adjustbox}
\newline
Table S2

### 2018-09-12 manuscript/supplement_methods.tex
final revisions edits from shawn before sending to co-auths

{\Large Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline

@@

The four LTER datasets each had different protocols for recording phenology observations. Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network. As in the NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation. The years used for each datasets were all years available at the time of analysis.

\textbf{Harvard Forest} \newline
We used observations from 1990-2014. Observations here were recorded as relative percentage of flowering or budburst for individual plants. We set "Yes" observations budburst and flower to the DOY when the percentage of each tree had was greater than or equal to 10\%

\textbf{H.J. Andrews Experimental Forest} \newline
We used observations from 2009-2015. We set "yes" observations for budburst to the first DOY when an individual was marked as "bud break" and "yes" observations for flowering were when an indiviudal was first marked as "Flowers open". Note that each species has slightly different ordinal codes to mark each of these events.

\textbf{Jornada Experimental Range} \newline
We used observations from 1992-2009. Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase. We set "Yes" observations for flowers to the first DOY where the flower phenophase was 10\% or greater.

@@

We set "Yes" observations for budburst to the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2].

>>>>

{\Comparison of large-scale citizen science data and long-term study data for phenology modeling \par} \newline

@@

The four LTER datasets each had different protocols for recording phenology observations. Below are details for converting the data from each to a status based yes/no as used in the National Phenology Network. As in the USA-NPN datasets, the julian day of year (DOY) used in modeling was the midpoint between each "yes" observation and the most recent "no" observation. The years used for each datasets were all years available at the time of analysis.

\textbf{Harvard Forest} \newline
We used observations from 1990-2014. Observations here were recorded as relative percentage of flowering or budburst for individual plants. We set "Yes" observations budburst and flower to the DOY when the percentage of each tree had was greater than or equal to 10\%. Harvard Forest has a sampling interval of 3-7 days.

\textbf{H.J. Andrews Experimental Forest} \newline
We used observations from 2009-2015. We set "yes" observations for budburst to the first DOY when an individual was marked as "bud break" and "yes" observations for flowering were when an individual was first marked as "Flowers open". Note that each species has slightly different ordinal codes to mark each of these events. H.J. Andrews has a sampling interval of 7 days.

\textbf{Jornada Experimental Range} \newline
We used observations from 1992-2009. Observations for this dataset represent, for each zone, the percent of plants for a particular species which is observed within each phenophase. We set "Yes" observations for flowers to the first DOY where the flower phenophase was 10\% or greater. Jornada has a sampling interval of 30 days.

@@

We set "Yes" observations for budburst to the DOY when the average value was greater than or equal to 1.6. This is the value where the 3 individuals most likely have ordinal observations of [1,2,2]. Hubbard Brook has a sampling interval of 7 days.
